---
title: "Building and testing SpaceRAT scaffolds for colorectal cancer consensus molecular subtypes"
author: "Puck Quarles van Ufford"
format: html
---

```{r}
#| label: seed

set.seed(8)
```

## Packages

```{r}
#| message: false
#| label: libraries

library("tidyverse")
library("spaceRAT")
```

## Load data

```{r}
#| message: false
#| label: load_data

TCGA_exprs <- read_delim("../data/04_TCGA_mapped.tsv")

NS_exprs <- read_delim("../data/04_NS_mapped.tsv")

cms_labels <- read_delim("../data/02_cms_labels_clean.tsv")
```

Reformat the expression data:

```{r}
#| label: reformat_data

TCGA_exprs <- TCGA_exprs |> 
  column_to_rownames("entrez")

NS_exprs <- NS_exprs |> 
  column_to_rownames("entrez")

cms_labels <- cms_labels |> 
  column_to_rownames("sample")
```

## Split data

We used 90% of the data to build the scaffold and the remaining 10% to project into the scaffold for testing.

```{r}
#| label: split_data
#| warning: false

samples <- rownames(cms_labels)

n <- length(samples)
test_set <- sample(seq_len(n), size = 0.1 * n)

TCGA_train <- TCGA_exprs |> 
  select(-test_set)
TCGA_test <- TCGA_exprs |> 
  select(test_set)

NS_train <- NS_exprs |> 
  select(-test_set)
NS_test <- NS_exprs |> 
  select(test_set)

cms_labels_train <- cms_labels |> 
  slice(-test_set)
cms_labels_test <- cms_labels |> 
  slice(test_set)
```

## SpaceRAT scaffolds for colorectal cancer

### TCGA data

Build the scaffold:

```{r}
#| label: build_TCGA_scafffold
#| message: false
#| warning: false

TCGA_scaffold <- buildScaffold(
  object = TCGA_train,
  pheno = cms_labels_train,
  colname = "cms",
  data = "exprs",
  annotation = "entrez"
)

dim(TCGA_scaffold$rank)
```

Plot the scaffold in PCA space:

```{r}
#| label: plot_TCGA_scafffold
#| message: false

plotScaffold(
  scaffold = TCGA_scaffold, 
  title = "CRC Scaffold built with RNA-seq data",
  dimred = "PCA", 
  dims = c(1,2), 
  plot_mode = "dot"
)

ggsave("../results/10_TCGA_scaffold.png")
```

Test the scaffold:

```{r}
#| label: test_TCGA_scafffold
#| message: false

projectSample(
  scaffold = TCGA_scaffold,
  sample = TCGA_test,
  pheno = cms_labels_test,
  colname = "cms",
  title = "CRC Scaffold built with RNA-seq data",
  annotation = NULL
)

ggsave("../results/10_TCGA_projections.png")
```

### NanoString data

Build the scaffold:

```{r}
#| label: build_NS_scafffold
#| message: false
#| warning: false

NS_scaffold <- buildScaffold(
  object = NS_train,
  pheno = cms_labels_train,
  colname = "cms",
  data = "exprs",
  annotation = "entrez"
)

dim(NS_scaffold$rank)
```

Plot the  scaffold in PCA space:

```{r}
#| label: plot_NS_scafffold
#| message: false

plotScaffold(
  scaffold = NS_scaffold, 
  title = "CRC Scaffold built with simulated NanoString nCounter data",
  dimred = "PCA", 
  dims = c(1,2), 
  plot_mode = "dot"
)

ggsave("../results/10_NS_scaffold.png")
```

Test the scaffold:

```{r}
#| label: test_NS_scafffold
#| message: false

projectSample(
  scaffold = NS_scaffold,
  sample = NS_test,
  pheno = cms_labels_test,
  colname = "cms",
  title = "CRC Scaffold built with simulated NanoString nCounter data",
  annotation = NULL
)

ggsave("../results/10_NS_projections.png")
```
